---
title: "Coral_P2_Week4Lab"
author: "`MIKO LIM YU HUI`, `TOH ZHENG YAN`, `LIM LIANG FAN`, `LIM JING CHUAN JONATHAN`, `HUANG WAN YING`, `TEE YU CHENG`, `ONG JIA EN DARRYL`"
format: html
date: "27 May 2025"
editor: visual
---

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# 1 Visualizing the Distribution of Travelers on the Titanic

## 1.1 Bar Chart of Travelers by Class

```{r}
#| label: bar-chart-class
titanic <- read_csv("data/titanic.csv")
titanic_count <- titanic %>% count(class)

ggplot(titanic_count, aes(x = class, y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), hjust = -0.1) +
  coord_flip() +
  labs(
    title = "Number of Travelers by Class",
    caption = "Source: Encyclopedia Titanica (2023)"
  ) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank())
```

## 1.2 Dodged Bar Chart of Travelers by Class and Survival

```{r}
#| label: dodged-bar-chart

titanic_survived <- titanic %>% count(class, survived)

ggplot(titanic_survived, aes(x = class, y = n, fill = survived)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_text(aes(label = n, color = survived),  # use 'color' aesthetic mapped to 'survived'
            position = position_dodge(width = 0.8), 
            hjust = -0.1) +
  coord_flip() +
  labs(
    title = "Survival by Class",
    fill = "Survived",
    color = "Survived"
  ) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank())

```

## 1.3 Mosaic Plot (Alternative: Heatmap with Matching Label Colors)

```{r}
#| label: heatmap-alternative

heatmap_data <- titanic %>% count(class, survived)

# Count the number of people per class and survival
titanic_counts <- titanic %>%
  count(class, survived)

# Compute total count
total_passengers <- sum(titanic_counts$n)

# Calculate the width (x) ranges per class
class_props <- titanic_counts %>%
  group_by(class) %>%
  summarise(class_total = sum(n)) %>%
  mutate(
    xmin = lag(cumsum(class_total / total_passengers), default = 0),
    xmax = cumsum(class_total / total_passengers)
  )

# Join back to main data and compute height (y) ranges
mosaic_data <- titanic_counts %>%
  left_join(class_props, by = "class") %>%
  group_by(class) %>%
  mutate(
    proportion = n / sum(n),
    ymin = lag(cumsum(proportion), default = 0),
    ymax = cumsum(proportion)
  )

# Create the plot
ggplot(mosaic_data) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = survived), color = "white") +
  geom_text(
    aes(x = (xmin + xmax) / 2, y = (ymin + ymax) / 2, label = n),
    color = "black", fontface = "bold"
  ) +
  scale_x_continuous(
    breaks = (class_props$xmin + class_props$xmax) / 2,
    labels = class_props$class
  ) +
  scale_y_continuous(
    breaks = c(0.25, 0.75),
    labels = c("FALSE", "TRUE")
  ) +
  labs(
    title = "Titanic Travelers by Class and Survival",
    x = "Class",
    y = "Survived",
    caption = "Source: Encyclopedia Titanica (2023)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## 1.4 Pros and Cons of Mosaic Plots

**Advantage:** Mosaic plots provide a clear visual representation of the relative proportions of each subgroup in a contingency table. In the Titanic example, the mosaic layout simultaneously shows both the total size of each class and the proportion of survivors and non-survivors within each class. This makes it easier to compare survival rates across different passenger classes in terms of both absolute and relative frequencies.

**Disadvantage:** Mosaic plots can be harder to interpret for audiences unfamiliar with them. The irregular tile sizes and dual-axis encoding (area = count) are less intuitive than standard bar charts. Also, exact values and differences between small subgroups may be visually ambiguous without labels, and mosaic plots are less effective when you want to emphasize the raw numbers rather than proportional relationships.

## 1.5 Alternative Visualization (Alluvial Plot)

```{r}
#| label: alluvial_stacked-bar-chart-alternative

library(ggalluvial)
library(dplyr)

# Prepare data
titanic_alluvial <- titanic %>%
  count(class, survived) %>%
  mutate(label = n)

# Alluvial plot
ggplot(titanic_alluvial, aes(axis1 = class, axis2 = survived, y = n)) +
  geom_alluvium(aes(fill = survived), width = 1/12, alpha = 0.8) +
  geom_stratum(width = 1/12, fill = "grey90", color = "black") +
  geom_text(
    stat = "stratum",
    aes(label = after_stat(stratum)),
    size = 4,
    check_overlap = TRUE
  ) +
  scale_x_discrete(limits = c("Class", "Survived"), expand = c(.05, .05)) +
  scale_fill_manual(values = c("TRUE" = "#66c2a5", "FALSE" = "#fc8d62")) +
  labs(
    title = "Passenger Flow from Class to Survival Outcome",
    y = "Number of Passengers",
    fill = "Survived"
  ) +
  theme_minimal()

```
**Advantages**:

-   Shows both **category sizes** and **directional relationships** in one coherent flow.

-   Makes it easier to trace **how different classes contributed to each survival outcome**, something that segmented or stacked bar charts can’t show as clearly.

<!-- -->

-    More engaging and **visually appealing**, which can improve audience attention in presentations.

**Disadvantages:**

-   **Harder to read exact quantities** without labels; thickness of ribbons can be misleading if not annotated properly.

-   Can be **more challenging to construct and interpret**, especially for audiences unfamiliar with this plot type.

-   Bar charts are better at showing **precise comparisons** between individual group sizes, while mosaic plots better emphasize **proportions within groups**.

# 2 Find a Data Set to Demonstrate Semantically Correct Aesthetic Mappings

## a. Provide the data URL
The dataset used in this analysis is publicly available from Our World in Data:

[owid-covid-data.csv](https://raw.githubusercontent.com/owid/covid-19-data/refs/heads/master/public/data/owid-covid-data.csv)

## b. Briefly summarize the scope, origin, and key variables of your data set in your own words

This dataset provides comprehensive daily COVID-19 data for countries and regions worldwide, sourced from official government and health ministry reports. It includes key variables such as the total number of COVID-19 cases per million people, total vaccinations administered per hundred people, population size, and continent information. The data is continually updated to reflect the evolving pandemic situation.

## c. Use ggplot2 (and any necessary add-on package) to produce a single plot incorporating at least two out of three aesthetic mapping categories listed in Section 2.1. Choose any plot type that best illustrates your variables.

```{r}
#| label: load-dataset

url <- "https://raw.githubusercontent.com/owid/covid-19-data/refs/heads/master/public/data/owid-covid-data.csv"
covid_data <- read_csv(url)

# Filter to most recent date per country
covid_latest <- covid_data %>%
  group_by(location) %>%
  # Arrange by descending date so first non-NA row will be the most recent valid one
  arrange(desc(date)) %>%
  # Find the first row with no NAs in all relevant columns
  filter(
    !is.na(continent),
    !is.na(total_cases_per_million),
    !is.na(total_vaccinations_per_hundred),
    !is.na(population)
  ) %>%
  slice(1) %>%
  ungroup()

# Select relevant columns and filter out NA values
covid_plot_data <- covid_latest %>%
  select(location, continent, population, total_cases_per_million, total_vaccinations_per_hundred) 

glimpse(covid_plot_data)
```

```{r}
#| label: covid-plot

ggplot(covid_latest, aes(
  x = total_vaccinations_per_hundred,
  y = total_cases_per_million,
  color = continent,
  size = population
)) +
  geom_point(alpha = 0.7) +
  scale_size(range = c(1, 12), guide = "legend") +
  labs(
    title = "COVID-19: Vaccinations vs Cases by Country",
    x = "Total Vaccinations per 100 people",
    y = "Total COVID-19 Cases per Million",
    color = "Continent",
    size = "Population"
  ) +
  theme_minimal()
```

## d. For each aesthetic mapping, explain why it is appropriate, referencing the variable nature and perceptual best practices.
* Color (continent): The continent variable is categorical, representing geographic regions. Using color to differentiate continents is appropriate because color is a highly effective way to visually distinguish categories, making it easier to compare regional patterns.
	
* Size (population): Population is an extensive numeric variable indicating country size. Mapping population to point size leverages the perceptual ability to compare magnitudes through area, effectively conveying how population scale relates to vaccination and case numbers.

## e. State the main insight or conclusion the reader should draw from your visualization.
The plot reveals a general trend that countries with higher vaccination rates tend to have varied case numbers, but continent-specific clusters appear—suggesting regional differences in pandemic dynamics and vaccine rollout. Additionally, larger countries (indicated by point size) show diverse pandemic outcomes, highlighting the complex relationship between population size, case counts, and vaccination coverage.

# 3 Reflections

## 1. Key takeaway: What is one thing you learned in today’s lab? Explain why this point stood out to you
We learned how to effectively combine multiple aesthetic mappings (color and size) in a single ggplot2 visualization to reveal complex relationships in real-world data. This stood out because it demonstrated how to encode multiple dimensions visually while maintaining clarity.

## 2. Challenges: What did you find confusing or difficult?
Filtering the dataset to include only the latest, complete records per country was tricky, especially handling missing values in a time series dataset. Understanding how to clean the data for accurate plotting took some effort.

## 3. AI assistance: If you used any generative AI tools (e.g., ChatGPT or GitHub Copilot), explain how they helped
We used ChatGPT to help us better understand the rationale behind mapping population size to point size and continent to color in our plot. The AI helped clarify the perceptual and statistical reasoning for these choices and provided guidance on how to explain these concepts clearly in our assignment.

## 4. Teamwork: How did your team collaborate on this assignment? Identify one strength and one area for improvement
Our team communicated well, dividing tasks like data sourcing and coding, which made the work efficient. One area to improve is better coordination on troubleshooting data issues earlier in the process.

##5. Additional feedback (optional): Any other comments on this module or the exercise
This exercise was valuable for applying data wrangling and visualization skills in RStudio with Quarto
