---
title: "Week 3 Lab: Exploring ggplot2"
author: "Coral Team"
format: html
date: "2025-05-20"
editor: visual
---

# Visualizing the Distribution of Quantitative Data: Bill Lengths of Pygoscelis Penguins

```{r}
#| label: load-libraries
#| message: false
library(tidyverse)
library(palmerpenguins)
library(ggplot2)
```

```{r}
#| label: load-data
#| message: false
data("penguins", package = "palmerpenguins")
```

## 1.1 Remove Missing Values

```{r}
#| label: remove-missing
penguins_clean <- drop_na(penguins, bill_length_mm)
```

## 1.2 Scatter Plot of Species versus Bill Length

```{r}
#| label: scatter-plot
#| echo: false
#| fig.width: 10
#| fig.height: 6

ggplot(penguins_clean, aes(x = bill_length_mm, y = species)) +
  geom_point() +
  labs(title = "Scatter Plot of Species vs. Bill Length",
       x = "Bill Length (mm)",
       y = "Species",
       caption = "Gorman et al. (2014)")
```

## 1.3 Why Should You Jitter the Plot?
Many penguins share the same species and have similar or identical bill length values, which leads to overplotting. Jittering helps distinguish overlapping points, improving visibility and interpretation.

## 1.4 Jitter Plot
```{r}
ggplot(penguins_clean, aes(x = bill_length_mm, y = species)) +
  geom_jitter(alpha = 0.6, width = 0, height = 0.2) +
  labs(
    title = "Jitter Plot of Bill Length by Species",
    x = "Bill Length (mm)",
    y = "Species",
    caption = "Source: Gorman et al. (2014)"
  )
```

## 1.5 Bee-Swarm Plot

## 1.6 Pros and Cons of Bee-Swarm Plots

## 1.7 Violin Plot with Boxes

## 1.8 What Does a Violin Plot Add to a Box Plot?

## 1.9 Frequency Polygons with Rug Plot

## 1.10 Faceted Frequency Polygons with Rug Plots

## 1.11 Pros and Cons of Faceting

## 1.12 Unmarked Bonus Question: Pairwise ð‘¡-Test
