---
title: "Week 3 Lab: Exploring ggplot2"
author: "Coral Team"
format: html
date: "2025-05-20"
editor: visual
---

# Visualizing the Distribution of Quantitative Data: Bill Lengths of Pygoscelis Penguins
```{r}
#| label: load-libraries
#| message: false
library(tidyverse)
library(palmerpenguins)
library(ggplot2)
```

```{r}
#| label: load-data
#| message: false
data("penguins", package = "palmerpenguins")
```

## 1.1 Remove Missing Values
```{r}
#| label: remove-missing
penguins_clean <- drop_na(penguins, bill_length_mm)
```


## 1.2 Scatter Plot of Species versus Bill Length 
```{r}
#| label: scatter-plot
#| echo: false
#| fig.width: 10
#| fig.height: 6

ggplot(penguins_clean, aes(x = bill_length_mm, y = species)) +
  geom_point() +
  labs(title = "Scatter Plot of Species vs. Bill Length",
       x = "Bill Length (mm)",
       y = "Species",
       caption = "Gorman et al. (2014)")
```

## 1.3 Why Should You Jitter the Plot?

## 1.4 Jitter Plot

## 1.5 Bee-Swarm Plot

## 1.6 Pros and Cons of Bee-Swarm Plots

## 1.7 Violin Plot with Boxes

## 1.8 What Does a Violin Plot Add to a Box Plot?

## 1.9 Frequency Polygons with Rug Plot

## 1.10 Faceted Frequency Polygons with Rug Plots

## 1.11 Pros and Cons of Faceting

## 1.12 Unmarked Bonus Question: Pairwise ð‘¡-Test
